Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-30 report/chapter04.Rnw
.

\section{Interpretation of Results}\label{sec:interpretation-results}
This section interprets the empirical evidence under the unified transport frame. We focus on TRTF (additive predictor), TTM-Sep, and, where applicable, copula baselines (only for $K\!\le\!3$) evaluated with matched preprocessing, metrics, and units. Synthetic studies report NLL, real datasets report LL, and we apply the shared affine correction. These commitments keep objectives, diagnostics, and compute interoperable across estimators.

TRTF often leads within the separable family because additive predictors shift conditional location while the underlying monotone shapes remain stable. The likelihood identities equate TRTF with separable triangular maps, so observed gaps arise from how each estimator realizes context shifts and stabilizes derivatives. On Half-Moon ($K=2$), TRTF achieved an NLL of $1.71$ while TTM-Sep reached $1.93$, and the first coordinate remained the main source of residual error. Table~\ref{tab:halfmoon-nll} records the per-dimension decomposition and associated uncertainty bands, showing that location adjustments dominate the remaining discrepancies when separability holds approximately in low dimensions.

@@

Performance on MINIBOONE reveals the cost of separability at higher dimension. TRTF improved the Gaussian reference yet remained about $18$~nats behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates. POWER exhibited the opposite regime: under identical preprocessing, the reported flows outperformed TRTF (Table~\ref{tab:uci-loglik} lists TRTF at $-7.17$ versus flow baselines near $0.17$ to $0.30$). These contrasts suggest that conditional shape and dimension jointly determine whether separable structure suffices. Table~\ref{tab:uci-loglik} reports these comparisons in a common unit.

@@

The outlook follows directly from the evidence. Data-driven orderings are likely to reduce anisotropy without abandoning the lower-triangular map. Low-rank cross-terms in triangular transports and non-additive predictors in TRTF may adapt conditional shapes while preserving monotone structure, exact inversion, and linear per-sample evaluation. We excluded these richer variants by design in Chapter~\ref{ch:intro} (Non-goals) due to compute and calibration overhead; they remain promising future work once resources permit. Expanded calibration reporting, including probability-integral-transform summaries and Kolmogorov--Smirnov distances, should remain part of any deployment-grade evaluation. Completing GAS and HEPMASS under the same protocol will improve generality and sharpen the accuracy-versus-compute trade-off. These steps target smaller likelihood gaps on high-$K$ datasets while retaining the interpretability and reproducibility provided by the transport frame.

>>>>

\section{Interpretation of Results}\label{sec:interpretation-results}
This section interprets the empirical evidence under the unified transport frame. We focus on TRTF, TTM-Sep, and, where applicable, copula baselines (only for $K\!\le\!3$) evaluated with matched preprocessing, metrics, and units. Synthetic studies report NLL, real datasets report LL, and we apply the shared affine correction. These commitments keep objectives, diagnostics, and compute interoperable across estimators.

TRTF often leads within the separable family because the forest aggregation shifts conditional location while the underlying monotone shapes remain stable. The likelihood identities equate TRTF with separable triangular maps, so observed gaps arise from how each estimator realizes context shifts and stabilizes derivatives. On Half-Moon ($K=2$), TRTF achieved an NLL of $1.71$ while TTM-Sep reached $1.93$, and the first coordinate remained the main source of residual error. Table~\ref{tab:halfmoon-nll} records the per-dimension decomposition and associated uncertainty bands, showing that location adjustments dominate the remaining discrepancies when separability holds approximately in low dimensions.

@@

Performance on MINIBOONE reveals the cost of separability at higher dimension. TRTF improved the Gaussian reference yet remained about $18$~nats behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates (TRTF recorded $-30.01$ while the strongest flow reached $-11.68$). POWER showed the complementary regime: under identical preprocessing, TRTF remains below the flow baselines (cf. Table~\ref{tab:uci-loglik}, where TRTF records $-7.17$ against flow baselines near $0.17$ to $0.30$). This contrast indicates that the gap is driven by conditional shape differences rather than pure location shifts. Table~\ref{tab:uci-loglik} reports these comparisons in a common unit.

@@

The outlook follows directly from the evidence. Data-driven orderings are likely to reduce anisotropy without abandoning the lower-triangular map. Low-rank cross-terms in triangular transports and non-separable predictors in TRTF may adapt conditional shapes while preserving monotone structure, exact inversion, and linear per-sample evaluation. We excluded these richer variants by design in Chapter~\ref{ch:intro} (Non-goals) due to compute and calibration overhead; they remain promising future work once resources permit. Completing GAS and HEPMASS under the same protocol will improve generality and sharpen the accuracy-versus-compute trade-off. These steps target smaller likelihood gaps on high-$K$ datasets while retaining the interpretability and reproducibility provided by the transport frame.

### 2025-09-30 report/chapterA1.Rnw
.

Maybe some R code here, probably a \rr{sessionInfo()}

>>>>

\section{Pseudo-code Summaries for Model Routines}

This appendix records consolidated pseudo-code for the core R implementations used in the experiments. Each summary captures inputs, main processing stages, and outputs so the execution flow is transparent without consulting the source code files.

\subsection{Transformation Random Forest (TRTF)}\label{app:trtf}

\textbf{Routine:}\quad\texttt{fit\_TRTF(S, config, seed, cores)} (calls \texttt{mytrtf}).

\begin{enumerate}
  \item Validate that the training matrix is numeric, set the RNG seed, and label columns as $X_1,\ldots,X_K$.
  \item Fit an intercept-only transformation model \texttt{BoxCox} for each $X_k$ to provide baseline monotone transformations.
  \item For $k = 2,\ldots,K$:
    \begin{enumerate}
      \item Build the formula $X_k \sim X_1 + \cdots + X_{k-1}$.
      \item Choose \texttt{mtry = max(1, floor((k-1)/2))} and standard \texttt{ctree} controls (\texttt{minsplit}, \texttt{minbucket}, \texttt{maxdepth}).
      \item Fit a transformation forest with \texttt{traforest} and store the conditional model (one forest per $k$).
    \end{enumerate}
  \item Return a \texttt{mytrtf} object containing baseline transformations, conditional forests, variable-importance scores, and the seed.

  \item \textbf{Prediction (\texttt{predict.mytrtf}):}
    \begin{enumerate}
      \item Convert new data to the same column naming scheme and evaluate $X_1$ through its baseline transformation model to obtain marginal log densities.
      \item For each conditional forest ($k\geq 2$) evaluate the log density of $X_k$ given $X_{1:(k-1)}$, extracting the diagonal when the forest returns a log density matrix.
      \item Stack the per-dimension log densities (\texttt{logdensity\_by\_dim}) or sum them to obtain the joint log likelihood (\texttt{logdensity}).
    \end{enumerate}
\end{enumerate}

\subsection{Nonparametric Copula Baseline}\label{app:copula}

\textbf{Routine:}\quad\texttt{fit\_copula\_np(S, seed)}.

\begin{enumerate}
  \item Inspect the training matrix and optional class labels; detect whether the dedicated copula packages are available.
  \item If prerequisites fail (dimension $K \neq 2$ or labels missing), fall back to independent univariate kernel density estimates per dimension and store them for later interpolation.
  \item Otherwise, for each class label:
    \begin{enumerate}
      \item Fit one-dimensional \texttt{kde1d} models to each marginal $X_1$ and $X_2$.
      \item Convert training samples to pseudo-observations using mid-ranks scaled by $(n+1)^{-1}$ and clamp to $(\varepsilon, 1-\varepsilon)$.
      \item Fit a two-dimensional kernel copula with \texttt{kdecopula::kdecop} (method \texttt{TLL2}).
      \item Store marginals, copula fit, and effective sample size for the class.
    \end{enumerate}
  \item Record class priors and return a \texttt{copula\_np} object.

  \item \textbf{Prediction (\texttt{predict.copula\_np}):}
    \begin{enumerate}
      \item In fallback mode evaluate each univariate KDE at the requested points and sum log densities.
      \item In copula mode compute marginal log densities and CDF values, evaluate the copula density, and either:
        \begin{enumerate}
          \item Average over class-specific log densities weighted by priors (mixture prediction), or
          \item Use the class labels supplied at prediction time.
        \end{enumerate}
      \item Return per-dimension log densities or their sum depending on the requested type.
    \end{enumerate}
\end{enumerate}

\subsection{Triangular Transport Core Utilities}\label{app:ttm-core}

\textbf{Module:}\quad\texttt{ttm\_core.R} (shared by marginal and separable TTM fits).

\begin{enumerate}
  \item Provide train-only standardization helpers that cache feature means and standard deviations and reapply them to new data.
  \item Define basis builders: polynomial features for predecessor coordinates $g_k$, monotone basis functions $f_k$ for the current coordinate, and their derivatives.
  \item Implement optional ordering heuristics (identity or Cholesky pivoting with optional Gaussianization) and persist selected permutations.
  \item Expose a dispatcher \texttt{ttm\_forward(model, X)} that:
    \begin{enumerate}
      \item Standardizes inputs using stored parameters.
      \item For marginal maps apply affine transformations $a_k + b_k x_k$ with precomputed coefficients.
      \item For separable maps constructs $g_k$ and $f_k$, computes $S_k = g_k + f_k$, and records the Jacobian diagonal $\partial_{x_k} S_k$.
    \end{enumerate}
  \item Provide \texttt{ttm\_ld\_by\_dim} to combine the forward map with the Gaussian reference, yielding per-dimension log densities used by all TTM variants.
\end{enumerate}

\subsection{Marginal Triangular Transport Map}\label{app:ttm-marg}

\textbf{Routine:}\quad\texttt{fit\_ttm\_marginal(data, seed)}.

\begin{enumerate}
  \item Split data into train/test subsets if only a matrix is provided; otherwise accept a prepared list.
  \item Standardize training features and, for each dimension $k$, compute closed-form coefficients $(a_k, b_k)$ that minimize the Gaussian pullback objective subject to $b_k > 0$.
  \item Store model parameters (standardization, per-dimension coefficients, ordering) and time measurements.
  \item During prediction call \texttt{ttm\_forward} with the marginal coefficients and convert Jacobian diagonals to log densities via \texttt{ttm\_ld\_by\_dim}; aggregate per-dimension contributions when the joint log density is requested.
\end{enumerate}

\subsection{Separable Triangular Transport Map}\label{app:ttm-sep}

\textbf{Routine:}\quad\texttt{fit\_ttm\_separable(data, degree\_g, lambda, seed)}.

\begin{enumerate}
  \item Prepare train/test splits and standardize training features as in the marginal case.
  \item For each coordinate $k$:
    \begin{enumerate}
      \item Build polynomial features $g_k$ on previous coordinates (degree set by \texttt{degree\_g}).
      \item Build monotone basis functions $f_k$ on the current coordinate and their derivatives.
      \item If \texttt{degree\_g = 0}, use the marginal closed-form solution to recover affine parameters.
      \item Otherwise solve the regularized optimization problem
            $\min_c \frac{1}{2}\lVert (I - \Phi_{\text{non}} M)c \rVert^2 - \sum \log (B c) + \lambda\,\text{penalty}(c)$
            using \texttt{optim} with L-BFGS-B while enforcing positivity of the derivative.
      \item Store coefficients $c_{\text{non}}$ and $c_{\text{mon}}$ for the coordinate.
    \end{enumerate}
  \item Assemble the model list with standardization parameters, coefficients, and metadata; record training/prediction timings.
  \item At prediction time re-use \texttt{ttm\_forward} and \texttt{ttm\_ld\_by\_dim} to obtain per-dimension and joint log densities.
\end{enumerate}

\subsection{Evaluation Utilities}\label{app:evaluation}

\textbf{Module:}\quad\texttt{evaluation.R} (experiment orchestration).

\begin{enumerate}
  \item Define convenience helpers such as \texttt{stderr(x)} and \texttt{add\_sum\_row} for table post-processing.
  \item \texttt{prepare\_data(n, config, seed)} samples from the configured data-generating process, splits the sample into train/validation/test sets, and returns both the matrix of draws and the split structure.
  \item \texttt{fit\_models(S, config)} fits the oracle TRUE density and the TRTF baseline on a split, times their evaluations, and returns the fitted objects together with per-dimension log-likelihood arrays.
  \item \texttt{calc\_loglik\_tables(models, config, X\_te, ...)} aggregates negative log-likelihoods (nats) for TRUE (marginal and joint), TRTF, TTM, and separable TTM, formats the results with standard-error bands, appends a summary row, and renames columns for presentation.
  \item \texttt{eval\_halfmoon(mods, S, out\_csv)} ensures all requisite models are available (TRTF, TTM variants, copula baseline), evaluates them on the half-moon test split, computes joint and per-dimension negative log-likelihoods, and optionally persists the metrics as CSV artifacts.
\end{enumerate}
